---
title: "Building vectorized Shiny UI / Server actions"
format: html
editor: visual

vignette: >
  %\VignetteIndexEntry{Building}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

## Motivation

ActionButtons & actionLinks are great components to communicate user actions from the UI of a Shiny app to its server.

But when it comes to generate many similar actions – for example one for each row of a data.frame – it becomes tricky to have one listener / observer for each of them on server side.

Since the action itself is expected to trigger the same code but with, for example, different ids, one might want to setup a mechanism where different buttons or links will trigger the same listener / observer and pass the id of the row as value.

## Customized onClick events

By default, two actionButtons will target two different inputs:

```{r}
#| eval: false

library(shiny)

# -- on UI side
actionButton(inputId = "btn_1", label = "Do this with row 1")
actionButton(inputId = "btn_2", label = "Do this with row 2")


# -- on Server side
observeEvent(input$btn_1, {
  # do something with row 1
})


observeEvent(input$btn_2, {
  # do something with row 2
})

```

In this scenario, the id of the row is never passed / used because `input$btn_1` will just be incremented each time the user clicks on the button.

So the idea is to somehow bypass the incremental mechanism to send a specific value to a single observer.\
That is when onClick comes into action:

```{r}
#| eval: false

actionButton(inputId = "btn_1", 
             label = "Do this with row 1",
             onClick = paste0('Shiny.setInputValue(\"action\", 1, {priority: \"event\"})'))

actionButton(inputId = "btn_2", 
             label = "Do this with row 1",
             onClick = paste0('Shiny.setInputValue(\"action\", 2, {priority: \"event\"})'))


observeEvent(input$action, {
  
  # -- perform action of specific row
  target_row <- input$action
  
})

```

## Helper functions

To reduce the amount of redundant code to write & avoid mistakes, ktools provides 3 functions that support this pattern:

- onclick_event()

- action_link()

- input_decode()

The first one helps building the onClick event instruction, including namespace when working with Shiny modules. It's called inside the action_link() function but could be used for example to tune an `actionButton`.

The action_link() function is a vectorized wrapper around the Shiny::actionLink() function that also includes namespace capabilities.

The last one is meant to decode a specific pattern that can be used as inputId(s) for the actionLinks HTML tags, so that it can be sent to the observer that will decode the source (namespace), the action and a value (most probably the row id where to apply the action).

## Use case

As said earlier, the motivation is to generated many similar actions to target vectorized / row-wise data.\
Let's create a basic dataset:

```{r}
#| eval: false

df <- data.frame(id = 1:10,
                 value = runif(10))
```

From there, we want to create an actionLink for each row of the data.frame:

```{r}
#| eval: false
library(ktools)

# -- Send default tag id to the target input:
links <- action_link(id = df$id, label = "click", target = "on_click")

```

This will generate a list of actionLink components with unique id(s) based on a default pattern: `"action_link_1"`, `"action_link_2"`, ...

When clicked, those links will trigger `input$on_click`, passing the component's id.

One can decide to make the component id(s) more explicit by using the `pattern` argument:

```{r}
#| eval: false

# -- Send custom tag id to the target input: 
action_link(id = df$id, label = "click", target = "on_click", pattern = "do_this")
```

This time, `input$on_click` will receive the custom tag id `"do_this_1"`, `"do_this_2"`, ...\
Which opens to using the same input entry to manage multiple actions inside the same observer / listener.

The input_decode() function is here to assist:

```{r}
#| eval: false

input_decode("do_this_1")
```

It returns a named vector with shape c(namespace, action, value) or c(action, value) if no namespace has been used to identify the source or the action, the action itself and it's target (value is the row id).

The listener / observer will look like this:

```{r}
#| eval: false

observeEvent(input$on_click, {
  
  # -- decode input value
  event <- input_decode(input$on_click)
  
  # -- do something on target row (here set it's 'value' column to 1)
  if(event$action == "do_this")
    df[df$id == event$value, ]$value <- 1
  
})

```

By extension, a unique `observeEvent` could be used to trigger various actions like add or subtract 1 to a quantity or set it to zero.

## Working with Shiny modules

The `namespace` argument in `action_link()` needs to be used when working with Shiny module to ensure the uniqueness of the HTML tag ids across the app.

It will also wrap the `target` input into the namespace so that the value will go to the right entry (i.e. in the module scope).

By extension, one could target a namespace used in a different module to trigger cross module communication (although it's breaks the encapsulation [pattern](https://thekangaroofactory.github.io/communication-between-shiny-modules/communication-between-modules.html "See Mastering Communication Between Shiny Modules")).

